---
title: "Getting started with blaze"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with blaze}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

blaze adds strict, opt-in type checking to ordinary R code. You write a type with
a small vocabulary, and blaze enforces it at runtime. Nothing is a separate
language; you add types where you want a guarantee and leave the rest untouched.

```{r setup}
library(blaze)
```

## Checking a value

`check()` verifies a value against a type. On success it returns the value
invisibly; on failure it raises a precise error. `int(1)` is a length-1 integer.

```{r}
check(5L, int(1))
```

```{r, error = TRUE}
check("x", int(1))
```

Types compose with the native pipe. Refinements such as `between()` and
`nonneg()` add element-wise constraints, and `NA` is rejected unless you allow it
with `na_ok()`.

```{r}
check(c(0.2, 0.9), dbl() |> between(0, 1))
```

```{r, error = TRUE}
check(c(1, -3), dbl() |> nonneg())
```

## Typed functions

`fn()` is a drop-in for `function()`. Declare an argument as `name = <type>` to
type-check it; anything else is an ordinary, unchecked default. The final
argument is the body, and `.returns` types the result.

```{r}
add <- fn(x = int(1), y = int(1), .returns = int(1), x + y)
add(2L, 3L)
```

```{r, error = TRUE}
add(2L, "x")
```

## Data-frame schemas

A `frame()` is a type describing a data frame's columns; `where()` adds row or
cross-column invariants. Every problem is reported together.

```{r}
patients <- data.frame(
  id  = 1:3,
  age = c(30L, 45L, 17L),
  grp = factor(c("ctrl", "treat", "ctrl"))
)

Patient <- type(frame(
  id  = int() |> unique_vals(),
  age = int() |> between(0, 120),
  grp = fct(c("ctrl", "treat"))
) |> where(age >= 18))
```

```{r, error = TRUE}
check(patients, Patient)
```

A schema is itself a type, so it drops straight into a signature:

```{r}
mean_age <- fn(data = frame(age = int() |> nonneg()), mean(data$age))
mean_age(data.frame(age = c(20L, 40L)))
```

## The vocabulary

The terse names live only inside blaze's constructs, so they never clash with
other packages. List them with `blaze_types()`, and read about any of them with,
for example, `?int`.

```{r}
blaze_types()
```

To build a reusable type outside `fn()`/`check()`/`frame()`, wrap it in `type()`:

```{r}
Probability <- type(dbl(1) |> between(0, 1))
check(0.5, Probability)
```
